Phenotypic evolution is more constrained in simpler food webs

### 2019-01-11 manuscript.tex

\documentclass[11pt,]{article}
\usepackage{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Matthew A. Barbour\^{}\{1,2,\textbackslash{}ast\}; Christopher J. Greyson-Gaito\^{}\{1,3\}; Arezoo Sootodeh\^{}\{3\}; Brendan Locke\^{}\{4\}; Jordi Bascompte\^{}\{2\}},
            pdftitle={Phenotypic evolution is more constrained in simpler food webs},
            colorlinks=true,
            citecolor=blue,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

%%%%%%%%% Previous - likely useful though
%\usepackage{setspace}\doublespacing
%\usepackage{float}
%\let\origfigure\figure
%\let\endorigfigure\endfigure
%\renewenvironment{figure}[1][2] {
%    \expandafter\origfigure\expandafter[H]
%} {
%    \endorigfigure
%}

@@

Global change is simplifying the structure of ecological networks;
however, we are currently in a poor position to predict how these
simplified communities will affect the evolutionary potential of
remaining populations. Theory on adaptive landscapes provides a
framework for predicting how selection constrains phenotypic evolution,
but often treat the community context of evolving populations as a
``black box''. Here, we integrate ecological networks and adaptive
landscapes to examine how changes in food-web complexity shape
evolutionary constraints. We conducted a field experiment that
manipulated the diversity of insect parasitoids (food-web complexity)
that were able to impose selection on an insect herbivore. We then
measured herbivore survival as a function of three key phenotypic
traits. We found that more traits were under selection in simpler
vs.~more complex food webs. The adaptive landscape was more neutral in
complex food webs because different parasitoid species impose different
selection pressures, minimizing relative fitness differences among
phenotypes. Our results suggest that phenotypic evolution becomes more
constrained in simplified food webs. This indicates that the
simplification of ecological communities may constrain the adaptive
potential of remaining populations to future environmental change.

@@

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how community
context constrains evolutionary change \citep{Arnold1992}. Different
aspects of evolutionary constraints can be inferred by quantifying the
slope and curvature of the adaptive landscape. For example, the slope is
determined by directional selection gradients acting on each phenotypic
trait and influences the trajectory of evolutionary change
\citep{Lande1979, Arnold1992}. Evolutionary trajectories become more
constrained with an increase in the number of traits under selection, as
this diminishes the number of optimal solutions \citep{Arnold2003}. The
curvature of the adaptive landscape can also constrain evolution through
its indirect effect on genetic constraints
\citep{Arnold1992, Hansen2008}. Genetic constraints are largely governed
by a population's \textbf{G}-matrix ---the additive genetic variances
and covariances between traits \citep{Hansen2008}. In general, genetic
constraints will increase with the number of traits under directional or
stabilizing selection, as this will decrease the additive genetic
variance in those traits \citep{Hansen2008}. Genetic constraints may
also increase with the number of trait combinations under correlational
selection, as this type of selection decreases the evolutionary
independence of traits \citep{Hansen2008}. If we want to predict how
community context constrains evolutionary change, we must understand how
ecological networks shape the adaptive landscape.

Global change is simplifying the structure of ecological communities,
which may influence evolutionary constraints in a number of ways. For
example, in a multitrophic community or food web, if different consumers
impose directional selection on different traits of a shared resource,
then more complex food webs may constrain evolution by increasing the
number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more complex food webs. To examine these
different possibilities (among others), we conducted a field experiment
that manipulated the complexity of trophic interactions (hereafter
food-web complexity) associated with an abundant insect herbivore
(\emph{Iteomyia salicisverruca}; Family Cecidomyiidae)(fig.
\ref{fig:Conceptual}). The larvae of this herbivore induce tooth-shaped
galls when they feed on the developing leaves of willow trees
\citep[\emph{Salix} sp.,][]{Russo2006}. These galls protect larva from
attack by generalist predators (e.g.~ants, spiders), but they suffer
high mortality from egg and larval parasitoids \citep{Barbour2016}. We
manipulated food-web complexity by either excluding the guild of larval
parasitoids (simple food web) or allowing both egg and larval
parasitoids to impose selection on gall midge traits (complex food web;
fig. \ref{fig:Conceptual}). We then applied modern statistical methods
to quantify the effect of food-web complexity on the slope and curvature
of the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

Note that we ommitted the upper triangle of each matrix for clarity
since it is simply the reflection of the lower triangle. The sign of
diagonal terms of the curvature matrix dictate whether selection will
increase (+), decrease (-), or cause no change (0) in the additive
genetic variance of a trait. Similarly, any nonzero covariance terms
(off-diagonal) are indicative of selection for trait integration (i.e.,
less independence). Therefore, we can infer the indirect effects of
selection on genetic constraints by counting the number of negative
signs along the diagonal (decrease in additive genetic variance) and the
number of nonzero terms along the off-diagonal (trait integration) of
the curvature matrix.

@@

Our key finding was that phenotypic evolution in gall midges was more
constrained in simpler food webs. More traits contributed to the slope
of the adaptive landscape in the simpler food web, suggesting that the
trajectory of phenotypic evolution is more constrained. Similarly, the
simple food-web treatment indirectly increased the number of genetic
constraints, which could act to constrain the gall midge's adaptive
potential in the face of novel selection pressures. We also found
evidence of indirect selection pressures in more complex food webs,
suggesting that the adaptive landscape may be more dynamic in complex
food webs. Taken together, our study provides experimental evidence from
the field that the simplification of ecological communities may
constrain the adaptive potential of remaining populations.

All three phenotypic traits we examined experienced directional
selection in the simpler food web, indicating that there is an optimal
phenotype that maximizes larval survival (i.e.~large chamber diameter,
small clutch size, and avoidance of conspecifics). In contrast, we did
not observe selective constraints on clutch size and oviposition
preference in the more complex food web. This suggests that there are
many optimal phenotypes (adaptive peaks), or in this case, a flat
adaptive landscape where there are no fitness consequences for
phenotypic change in these traits. This also implies that as selective
constraints dampen in more complex food webs, then the trajectory of
evolution becomes more determined by genetic constraints. This is made
clear if we inspect the `Lande equation', \(\Delta z = \text{G} \beta\),
where G corresponds to the population's G-matrix. As the components of
\(\beta\) go to zero, then \(\Delta z\) is more influenced by the
G-matrix. Interestingly, this corresponds to \citet{Schluter1996}`s
hypothesis that phenotypic evolution often follows 'genetic lines of
least resistance'. \citet{Schluter1996} found support for this
hypothesis from data on natural populations of several vertebrate
species, including threespine sticklebacks, a few species of songbirds,
and mice from the genus \emph{Peromyscus}. All of these species occupy
intermediate trophic levels and are likely embedded in complex food
webs, which is consistent with our suggestion that genetic constraints
may have a stronger influence in more complex food webs.

There are a growing number of theoretical studies on adaptation to
directional environmental change that incorporate species interactions
\citep[e.g.][]{deMazancourt2008, Johansson2008, Norberg2012, Osmond2017PredatorsHelpPrey}.
This work has given insight to the mechanisms by which pairwise
interactions modify the adaptive potential of species. Our study hints
at a novel mechanism, whereby food-web complexity flattens the adaptive
landscape, thus facilitating future adaptation by allowing genetic and
phenotypic variation to persist. This mechanism only emerges once we
move beyond pairwise interactions to consider selection on multiple
traits in a community context. However, a study on competition has
highlighted that we may expect the opposite effect of species diversity
in competitive communities \citep{deMazancourt2008}. This negative
effect of species diversity on adaptation occurs because there is a
greater chance that the optimal phenotype is already occupied by a
competitor in a more diverse community. More work is needed to examine
how the distribution of different interaction types affects adaptation
to environmental change in species-rich communities.

\section{References}\label{references}

\hypertarget{refs}{}

@@

@@

\begin{figure}
\centering
\includegraphics{analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in complex vs.~simple food webs. Each panel corresponds to a
different phenotypic trait: chamber diameter (A); oviposition preference
(B); and clutch size (C). Solid lines represent selection experienced in
complex (orange) and simple (blue) food webs. Transparent lines
represent bootstrapped replicates to show the uncertainty in selection.
For clarity, we only display 100 bootstraps even though inferences are
based on 1,000 replicates. Note that mean larval survival is plotted on
a natural log scale to reflect the mathematical defintion of the
adaptive landscape.}
\end{figure}

\begin{figure}
\centering
\includegraphics{analyses/selection_on_Platygaster.pdf}
\caption{\label{fig:EggPtoid_Selection}Selection imposed by larval
parasitoids on the extended phenotype of egg parasitoids
(\emph{Platygaster} sp.). The solid line represents the average
difference in the probability of observing the egg parasitoid in complex
vs.~simple food webs as a function of gall midge oviposition preference.
Transparent lines represent bootstrapped replicates to show the
uncertainty in selection. For clarity, we only display 100 bootstraps
even though inferences are based on 1,000 replicates. The decrease in
the probability of observing egg parasitoids at high gall-midge
densities indicate that larval parasitoids impose nonlinear selection on
the extended phenotype of egg parasitoids.}
\end{figure}

>>>>

\documentclass[11pt,]{article}
\usepackage{mathpazo}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
    \usepackage{xltxtra,xunicode}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Mapping=tex-text,Scale=MatchLowercase}
  \newcommand{\euro}{€}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage{natbib}
\bibliographystyle{amnat}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\ifxetex
  \usepackage[setpagesize=false, % page size defined by xetex
              unicode=false, % unicode breaks when used with xetex
              xetex]{hyperref}
\else
  \usepackage[unicode=true]{hyperref}
\fi
\hypersetup{breaklinks=true,
            bookmarks=true,
            pdfauthor={Matthew A. Barbour\^{}\{1,2,\textbackslash{}ast\}; Christopher J. Greyson-Gaito\^{}\{1,3\}; Arezoo Sootodeh\^{}\{3\}; Brendan Locke\^{}\{4\}; Jordi Bascompte\^{}\{2\}},
            pdftitle={Phenotypic evolution is more constrained in simpler food webs},
            colorlinks=true,
            citecolor=black,
            urlcolor=black,
            linkcolor=black,
            pdfborder={0 0 0}}
\urlstyle{same}  % don't use monospace font for urls
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\setcounter{secnumdepth}{0}

@@

%%%%%%%%% Previous - likely useful though
%\usepackage{setspace}\doublespacing
\usepackage{float}
\let\origfigure\figure
\let\endorigfigure\endfigure
\renewenvironment{figure}[1][2] {
    \expandafter\origfigure\expandafter[H]
} {
    \endorigfigure
}

@@

Global change is simplifying the structure of ecological networks;
however, we are currently in a poor position to predict how these
simplified communities will affect the evolutionary potential of
remaining populations. Theory on adaptive landscapes provides a
framework for predicting how selection constrains phenotypic evolution,
but often treats the community context of evolving populations as a
``black box''. Here, we integrate ecological networks and adaptive
landscapes to examine how changes in food-web complexity shape
evolutionary constraints. We conducted a field experiment that
manipulated the diversity of insect parasitoids (food-web complexity)
that were able to impose selection on an insect herbivore. We then
measured herbivore survival as a function of three key phenotypic
traits. We found that more traits were under selection in simpler
vs.~more complex food webs. The adaptive landscape was more neutral in
complex food webs because different parasitoid species impose different
selection pressures, minimizing relative fitness differences among
phenotypes. Our results suggest that phenotypic evolution becomes more
constrained in simplified food webs. This indicates that the
simplification of ecological communities may constrain the adaptive
potential of remaining populations to future environmental change.

@@

Ecological networks, such as food webs describing who-eats-whom in
ecological communities, provide an explicit representation of the
community context \citep{Bascompte2014, McCann2012}. Here, we integrate
ecological networks and adaptive landscapes to understand how the
community context constrains evolutionary change \citep{Arnold1992}.
Different aspects of evolutionary constraints can be inferred by
quantifying the slope and curvature of the adaptive landscape. For
example, the slope is determined by directional selection gradients
acting on each phenotypic trait and influences the trajectory of
evolutionary change \citep{Lande1979, Arnold1992}. Evolutionary
trajectories become more constrained with an increase in the number of
traits under selection, as this diminishes the number of optimal
solutions \citep{Arnold2003}. The curvature of the adaptive landscape
can also constrain evolution through its indirect effect on genetic
constraints \citep{Arnold1992, Hansen2008}. Genetic constraints are
largely governed by a population's \textbf{G}-matrix ---the additive
genetic variances and covariances between traits \citep{Hansen2008}. In
general, genetic constraints will increase with the number of traits
under directional or stabilizing selection, as this will decrease the
additive genetic variance in those traits \citep{Hansen2008}. Genetic
constraints may also increase with the number of trait combinations
under correlational selection, as this type of selection decreases the
evolutionary independence of traits \citep{Hansen2008}. If we want to
predict how the community context constrains evolutionary change, we
must understand how ecological networks shape the adaptive landscape.

Global change is simplifying the structure of ecological communities,
which may influence evolutionary constraints in a number of ways. For
example, in a multitrophic community or food web, if different consumers
impose directional selection on different traits of a shared resource,
then more complex food webs may constrain evolution by increasing the
number of traits under selection. Alternatively, if consumers impose
selection on different values of a trait, then their selective effects
would cancel each other out in more complex food webs. To examine these
different possibilities, we conducted a field experiment that
manipulated the complexity of trophic interactions (hereafter food-web
complexity) associated with an abundant insect herbivore (\emph{Iteomyia
salicisverruca}, Family Cecidomyiidae; fig. \ref{fig:Conceptual}). The
larvae of this herbivore induce tooth-shaped galls when they feed on the
developing leaves of willow trees \citep[\emph{Salix}
sp.,][]{Russo2006}. These galls protect larva from attack by generalist
predators (e.g.~ants, spiders), but they suffer high mortality from egg
and larval parasitoids \citep{Barbour2016}. We manipulated food-web
complexity by either excluding the guild of larval parasitoids (simple
food web) or allowing both egg and larval parasitoids to impose
selection on gall midge traits (complex food web; fig.
\ref{fig:Conceptual}). We then applied modern statistical methods to
quantify the effect of food-web complexity on the slope and curvature of
the gall midge's adaptive landscape. Taken together, our study gives
insight to how local extinctions, and concomitant decreases in food-web
complexity, may constrain the evolution of interacting populations.

@@

Note that we ommitted the upper triangle of each matrix for clarity
since it is simply the reflection of the lower triangle. The sign of
diagonal terms of the curvature matrix dictates whether selection will
increase (+), decrease (-), or cause no change (0) in the additive
genetic variance of a trait. Similarly, any nonzero covariance terms
(off-diagonal) are indicative of selection for trait integration (i.e.,
less independence). Therefore, we can infer the indirect effects of
selection on genetic constraints by counting the number of negative
signs along the diagonal (decrease in additive genetic variance) and the
number of nonzero terms along the off-diagonal (trait integration) of
the curvature matrix.

@@

@@

\begin{figure}
\centering
\includegraphics{analyses/UV_landscapes.pdf}
\caption{\label{fig:UV_Landscape}Adaptive landscape of gall midge
phenotypes in complex vs.~simple food webs. Each panel corresponds to a
different phenotypic trait: chamber diameter (A); oviposition preference
(B); and clutch size (C). Bold lines represent selection experienced in
complex (orange) and simple (blue) food webs. Thin lines represent
bootstrapped replicates to show the uncertainty in selection. For
clarity, we only display 100 bootstraps even though inferences are based
on 1,000 replicates. Note that mean larval survival is plotted on a
natural log scale to reflect the mathematical defintion of the adaptive
landscape.}
\end{figure}

\begin{figure}
\centering
\includegraphics{analyses/selection_on_Platygaster.pdf}
\caption{\label{fig:EggPtoid_Selection}Selection imposed by larval
parasitoids on the extended phenotype of egg parasitoids
(\emph{Platygaster} sp.). The bold line represents the average
difference in the probability of observing the egg parasitoid in complex
vs.~simple food webs as a function of gall midge oviposition preference.
Thin lines represent bootstrapped replicates to show the uncertainty in
selection. For clarity, we only display 100 bootstraps even though
inferences are based on 1,000 replicates. The decrease in the
probability of observing egg parasitoids at high gall-midge densities
indicate that larval parasitoids impose nonlinear selection on the
extended phenotype of egg parasitoids.}
\end{figure}

We found that simpler food webs constrained phenotypic evolution in gall
midges in two key ways. First, more traits contributed to the slope of
the adaptive landscape in the simpler food web, suggesting greater
constraints on the trajectory of phenotypic evolution. Second, the
simple food-web treatment indirectly increased the number of genetic
constraints, which could act to constrain the gall midge's adaptive
potential in the face of novel selection pressures. We also found
evidence of indirect selection pressures in more complex food webs,
suggesting that the adaptive landscape may be more dynamic in complex
food webs. Taken together, our study provides experimental evidence from
the field that the simplification of ecological communities may
constrain the adaptive potential of remaining populations.

All three phenotypic traits we examined experienced directional
selection in the simpler food web, indicating that there is an optimal
phenotype that maximizes larval survival (i.e.~large chamber diameter,
small clutch size, and avoidance of conspecifics). In contrast, we did
not observe selective constraints on clutch size and oviposition
preference in the more complex food web. This suggests that there are
many optimal phenotypes (adaptive peaks), or in this case, a flat
adaptive landscape where there are no fitness consequences for
phenotypic change in these traits. This also implies that as selective
constraints dampen in more complex food webs, then the trajectory of
evolution becomes more determined by genetic constraints. This is made
clear if we inspect the `Lande equation', \(\Delta z = \text{G} \beta\),
where G corresponds to the population's G-matrix. As the components of
\(\beta\) go to zero, then \(\Delta z\) is more influenced by the
G-matrix. Interestingly, this aligns with \citet{Schluter1996}`s
hypothesis that phenotypic evolution often follows 'genetic lines of
least resistance'. \citet{Schluter1996} found support for this
hypothesis from data on natural populations of several vertebrate
species, including threespine sticklebacks, a few species of songbirds,
and mice from the genus \emph{Peromyscus}. All of these species occupy
intermediate trophic levels and are likely embedded in complex food
webs, which is consistent with our suggestion that genetic constraints
may have a stronger influence in more complex food webs.

There is a growing number of theoretical studies on adaptation to
directional environmental change that incorporate species interactions
\citep[e.g.][]{deMazancourt2008, Johansson2008, Norberg2012, Osmond2017PredatorsHelpPrey}.
This work has given insight to the mechanisms by which pairwise
interactions modify the adaptive potential of species. Our study hints
at a novel mechanism, whereby food-web complexity flattens the adaptive
landscape, thus facilitating future adaptation by allowing genetic and
phenotypic variation to persist. This mechanism only emerges once we
move beyond pairwise interactions to consider selection on multiple
traits in a community context. However, a study on competition has
highlighted that we may expect the opposite effect of species diversity
in competitive communities \citep{deMazancourt2008}. This negative
effect of species diversity on adaptation occurs because there is a
greater chance that the optimal phenotype is already occupied by a
competitor in a more diverse community. More work is needed to examine
how the distribution of different interaction types affects adaptation
to environmental change in species-rich communities.

@@
